Morse Theory

### 2020-12-04 paper.tex
POLISHED

\usepackage[utf8]{inputenc}
\usepackage{setspace}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{tikz}
\usetikzlibrary{matrix}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{proposition}{Proposition}
\newtheorem{scolium}{Scolium} 
\newtheorem{definition}{Definition}

@@

\begin{definition}(Morse function)
A smooth function on a manifold is a Morse function if all critical points are non-degenerate and all the critical points have distinct function values
\end{definition}
    \paragraph{\textbf{Morse Lemma.}}
        Suppose that the point $a \in \mathbb{R}^k$ is a nondegenerate
        critical point of the function $f$, and
        $$
            (h_{ij})= \left(  \frac{\partial^2f}{\partial x_i \partial x_j} (a) \right)
        $$
        is the Hessian of $f$ at $a$. Then there exists a local coordinate system $(x_1,\ldots,x_k)$ around $a$ such that
        $$
            f=f(a)+\sum{h_{ij}x_ix_j}
        $$
        near $a$.
        \\
        or
        $$
            f=-x_1^2-x_2^2-\ldots-x_\lambda^2+x_\lambda^2+\ldots+x_n^2+f(a)
        $$
        where $\lambda$ is the index of $f$ at $a$.

    \paragraph{\textbf{Theorem 1.1}} \textit{ 
        Let $f$ be a smooth function in a neighborhood 
    $N_x$ of $x=(x_1,\ldots,x_n)$ in $\mathbb{R}^n$. Suppose $f(0,\ldots,0)=0$. Then, 
    there exist $n$ smooth functions $g_i,\ldots,g_n$ defined on $N_x$ such that $g_i(0,\ldots,0)=\frac{\partial f}{\partial x_i}(0,\ldots,0)$
    for every $i$, and
    }

    $$
        f(x_i,\ldots,x_n)=\sum_{i=1}^{n}{(x_i,\ldots,x_n)}
    $$

    \paragraph{\textbf{Theorem 1.2}} \textit{ (Inverse Function Theorem)
        Let $f:\mathbb{R}^n\to \mathbb{R}^n$ be a smooth function on an open set $U$
        containing $a\in\mathbb{R}^n$. Suppose that $\det J_f(a)\not = 0$.
        \\Then there is an open set $V \subset \mathbb{R}^n$ containing $a$ and an open
        set $W\subset \mathbb{R}^n$ containing $f(a)$ such that $f:V\to W$ is a 
        diffeomorphism.
    }

    \paragraph{\textbf{Proof.}} Let $p_0$ be a nondegenerate critical point of the function
     $f:\ M \to \mathbb{R}$, where $M$ is an $n$-manifold. The degeneracy of the point
     $p_0$ on $f$ is determined independent of our choice of a local coordinate system.
     Therefore, we may assume that when we pick a local coordinate system $(x_1,\ldots,x_n)$ defined
     in a neighborhood $N_{p_0}$,

@@

     When we square $y_1$, we get
     \begin{equation}
        y_1^2=\pm h_{11}x_1^2\pm2\sum_{i=2}^{n}{x_1x_ih_{1i}}\pm\frac{\left(\sum_{i=2}^{n}{x_ih_{1i}}\right)^2}{h_{11}}\tag{1.9}
     \end{equation}
     where the signs are either positive or negative, depending on the sign of $h_{11}$. Using equation
     $1.5$, we can verify that $f$ can be expressed in the following way with respect to this
     coordinate system on the restricted domain $\tilde{N}_0$.
     \begin{equation}
        f=\pm y_1^2 + \sum_{i=2}^{n}\sum_{j=2}^{n}{x_ix_jh_{ij}}-\frac{\left(\sum_{i=2}^{n}{x_ih_{1i}}\right)^2}{h_{11}}\tag{1.10}
     \end{equation}
     where the sign of the $y_1^2$ term is positive or negative, depending on the sign of $h_{11}$. Staying
     consistent with our goals, we notice that the first term is in the standard quadratic form seen
     in the Morse Lemma formulation, whereas the rest of the terms depend on local coordinates
     $x_i$ whereby $i\not=k \ (k=1)$. By induction from $k=1$ to $k=n$, we prove the Morse Lemma. $\blacksquare$
% % end N1
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% % begin M1

\subsection{Proof that$f: S^2 \to \R, f(x,y,z) = z$ is a Morse function}
\textbf{Ex.} Show that the function $f: S^2 \to \R, f(x,y,z) = z$ is a Morse function. \\

$f$ is smooth on $ S^2 $ since it extends to a smooth map on all of $\R^3$. We can map $R^3$ to $S^2$ with the stereographic projection two functions:\\
$
	\phi_1(x_1,x_2,x_3) = (\frac{x_1}{1-x_3}, \frac{x_2}{1-x_3} ) 
$
and
$
	\phi_2(y_1,y_2,y_3) = (\frac{y_1}{1+y_3},\frac{y_2}{1+y_3})
$\\
Therefore, to get $S^2 \to R^3$ we can take $\phi_1^{-1}$ and $\phi_2^{-1}$ \\
$
	\phi_1^{-1}(x_1,x_2,x_3) = (\frac{2x_1}{x_1^2+x_2^2+1}, \frac{2x_2}{x_1^2+x_2^2+1}, \frac{x_1^2+x_2^2-1}{x_1^2+x_2^2+1})
$\\
$
	\phi_1^{-1}(y_1,y_2,y_3) = (\frac{2y_1}{y_1^2+y_2^2+1}, \frac{2y_2}{y_1^2+y_2^2+1}, \frac{1-y_1^2+y_2^2}{y_1^2+y_2^2+1})
$\\
Then, we take $g_1 = f \circ \phi_1^{-1}$ and $g_2 = f \circ \phi_2^{-1}$ \\
$
	g_1(x,y) =  \frac{x^2+y^2-1}{x^2+y^2+1}
$
and
$
	g_2(x,y) = \frac{1-x^2+y^2}{x^2+y^2+1}
$ \\
Now, we can compute jacobian of $g_1$ and $g_2$, find critical points, and check that determinant of hessian matrix is non-zero. \\
$
	\nabla g_1(x,y,z) = 0$ iff $(x,y,z) = (0,0,-1)
$\\
$
	\nabla g_2(x,y,z) = 0$ iff $ (x,y,z) = (0,0,1)
$
We have two critical points, and now we compute the hessian at them.\\

$
	H(g_1) = \begin{bmatrix} 4 & 0 \\ 0 & 4 \end{bmatrix}
	\imply \det(H(g_1)) = 16
$ \\

$
	H(g_2) = \begin{bmatrix} -4 & 0 \\ 0 & -4 \end{bmatrix}
	\imply \det(H(g_2)) =  16
$ \\

Both critical points are non-degenerate, therefore $f$ is a Morse function.$\blacksquare$\\
% % end M1
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% % start V1
\\
\\
\subsection{Examples of Morse functions} 
\textbf{Example 1.1}\\
$f(x,y)=e^{xy}+x$\\
First of all we have to find all critical points of $f$. A point $P$ is critical, when $\frac{\partial f}{\partial x}=\frac{\partial f}{\partial y}=0$ at $P$. \\
$\frac{\partial f}{\partial x}= ye^{xy}+1=0 \\
\frac{\partial f}{\partial y}=xe^{xy}=0$
After solving the system of equations we get that $x=0$, $y=-1$. Hence, the only critical point of $f$ is $(0,-1)$. Now to prove that the function is a Morse function we have to shoe that the determinant of Hessian matrix at the point $(0,-1)$ is non-zero. \\
Partial second order derivatives are: \\
$\frac{\partial^2 f}{\partial x^2}=y^2e^{xy}$
$\frac{\partial^2 f}{\partial x^2}(0,-1)=1$\\
$\frac{\partial^2 f}{\partial \partial y}=e^{xy}+yxe^{xy}$
$\frac{\partial^2 f}{\partial \partial y}(0,-1)=1$\\
$\frac{\partial^2 f}{\partial y \partial x}=e^{xy}+xye^{xy}$
$\frac{\partial^2 f}{\partial y \partial x}(0,-1)=1$\\
$\frac{\partial^2 f}{\partial y^2}=x^2e^{xy}$
$\frac{\partial^2 f}{\partial y^2}(0,-1)=0$\\
Therefore, \\
$det(Hess(f(0,-1)))=\left| \begin{array}{cc} 1 & 1 \\ 1 & 0 \end{array} \right|=-1$. \\
Since the determinant is non-zero at the only critical point of the function, we can claim that it is a Morse function. \\
\textbf{Example 1.2}\\
Let's consider a function $g(x,y)=xy$ and prove that it is , indeed, a Morse function.\\
First order partial derivatives: \\
$\frac{\partial g}{\partial x}=y$\\
$\frac{\partial g}{\partial y}=x$\\
Therefore, the only critical point of $g$ is $(0,0)$. Now we have to evaluate the determinant of Hessian matrix at this point.\\
Second order partial derivatives: \\
$\frac{\partial^2 g}{\partial x^2}=0$\\
$\frac{\partial^2 g}{\partial x \partial y}=1$\\
$\frac{\partial^2 g}{\partial y \partial x}=0$\\
$\frac{\partial^2 g}{\partial^2 y}=1$\\
Hence the determinant of Hessian at the critical point is: \\
$det(Hess(f(0,0)))=\left| \begin{array}{cc} 0 & 1 \\ 1 & 0 \end{array} \right|=-1$. \\
We observe that the determinant is nonzero, hence, the only critical point of $g(x,y)=xy$ is non-degenerate, so the function is Morse. \\
\\
\\
\subsection{Examples of non-Morse functions} 
\textbf{Example 2.1}\\
$f(x,y)=x^3+xY^2-x^2y-y^3$\\
Let's follow the previous procedure: \\
$\frac{\partial f}{\partial x}=3x^2+y^2-2xy=0$\\
$\frac{\partial f}{\partial y}=2xy-x^2-3y^2=0$\\
After solving the system of equations we got that the only critical point of $f$ is $(0,0)$. Now compute second order partial derivatives:\\
$\frac{\partial^2 f}{\partial x^2}=6x-2y$\\
$\frac{\partial^2 f}{\partial x \partial y}=2y-2x$\\
$\frac{\partial^2 f}{\partial y \partial x}=2y-2x$\\
$\frac{\partial^2 f}{\partial^2 y}=2x-6y$\\
Therefore, \\
$det(Hess(f(0,0)))=\left| \begin{array}{cc} 0 & 0 \\ 0 & 0 \end{array} \right|=0$.\\
Hence, the only critical point of $f$ is degenerate, so $f$ is not a Morse function. \\
$g(x,y)=x^3$\\
$\frac{\partial g}{\partial x}=3x^2$\\
$\frac{\partial g}{\partial y}=0$\\
Therefore, critical points are all points of the form $(0, y_1)$.\\
$\frac{\partial^2 g}{\partial x^2}=6x$
$\frac{\partial^2 g}{\partial x \partial y}=0$\\
$\frac{\partial^2 g}{\partial y \partial x}=0$
$\frac{\partial^2 g}{\partial^2 y}=0$\\
Hence, Hessian will be of the form: \\
$Hess(g(0,y_1))=\left| \begin{array}{cc} 0 & 0 \\ 0 & 0 \end{array} \right|,  \forall y_1$. \\
Therefore, all critical points of $g$ are degenerate points, moreover, values of $g$ at all critical points are equal, hence, $g(x,y)=x^3$ is not a Morse function. 
% % end V1
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\section{Reeb graphs}
     \subsection{Introduction to Reeb graphs}
     \paragraph{}
      Suppose we have an equivalence relation, $\sim$, defined on a topological space $\X$. Let $\X_\sim$
      be the set of equivalence classes and let $\phi : \X \to \X_\sim$ map each point $x$ to its equivalence class.

     \paragraph{\textbf{Definition 1.2.1}}
      Let $f: \X \to \R$ be continuous and call a component of a level set a contour. Two
      points $x,y \in \X$ are equivalent if they belong to the same component of $f^{-1}(t)$
      with $t=f(x)=f(y)$. The Reeb graph of $f$, denoted as $R(f)=\X_\sim$, is the quotient space 
      defined by this equivalence relation.

      By construction, the Reeb graph has a point for each contour and the connection is 
      provided by the map $\phi:\X\to R(f)$. Letting $\pi:R(f)\to\R$ be defined such that $f(x)=\pi(\phi(x))$
      we can construct the level set by going backward, from the real line to the Reeb graph to the 
      topological space. Given $t\in\R$ we get $\pi^{-1}(t)$, a collection of points in $R(f)$,
      and $\phi^{-1}(\pi^{-1}(t))$, the corresponding collection of contours that make up the level set defined by $t$.

@@

\section{Applications}
\subsection{Topological Data Analysis}
A computational method for extracting simple descriptions of high dimensional data sets is a Mapper algorithm:
Let $X$ be a point cloud (data set) and we have a function $f:X \to \R$.
Assume that we have $N$ points in $X$ whose outputs are known for the function $f$.
Also, we assume that it is possible to compute inter-point dinstances between the points in the data.
Let $I$ be the range of the function $f$ restricted to the $N$ points and let $S$ be the covering of $I$.
We begin by dividing $S$ into a smaller intervals $(S_i)$ which overlap.
This gives us two parameters $l_i$ the length of $S_i$ and $p_{ij}$ the perceptange overlap between successive intervals.
Now for each $S_i$ we find a set $X_j = \{x | f(x) \in S_j\}$ s.t. $X_j \subset X$. The set $\{X_j\}$ forms a cover of $X$, and $X \subset \cup_j X_j$.
For each smaller set $X_j$ we find clusters $\{X_{jk}\} $.
We treat each cluster as a vertex in our reeb graph and draw an edge between vertices whenever $X_{jk} \cap X_{lm} \neq \emptyset$ i.e. the clusters corresponding to vertices have non-empty intersection. \\

@@

We have implemented the Mapper algorithm and ran it for the sk learn diabetes dataset and standford cars dataset, output are as follows:

\begin{center}
\includegraphics[width=.4\textwidth]{diabetes.png}
\includegraphics[width=.4\textwidth]{cars.png}
\end{center}
Based on the diabetes data we may be able to make a difference between patients who don't need a treatmant, diabetis 1 and diabetis 2. While in the cars dataset it looks like there aren't any great differences between car features.

\textbf{Applications of Reeb Graphs in Computer Graphics}\\
Computer graphics, in particular 3D modeling and animation, are a vast field. Building 3D models and analyzing their features are important both in entertainment (cartoons, animated films, computer games) and for scientific research (building 3D models based on photographs of a microscope for biology and medicine; building physical interactions in mechanics, astronomy, atomic physics ). There are two problems associated with 3D modeling, which are solved using Reeb graphs.\\

@@

\begin{thebibliography}{9}
     \bibitem{s}
         {\sc }
         `` https://core.ac.uk/download/pdf/20670562.pdf
\end{thebibliography}
\end{document}

>>>>

\usepackage[utf8]{inputenc}
\usepackage{setspace}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{tikz}
\usetikzlibrary{matrix}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{example}{Example}
\newtheorem{problem}{Problem}
\newtheorem{proposition}{Proposition}
\newtheorem{scolium}{Scolium} 
\newtheorem{definition}{Definition}

@@

\begin{definition}(Morse function)
A smooth function on a manifold is a Morse function if 
\begin{itemize}
	\item all critical points are non-degenerate
	\item the critical points have distinct function values
\end{itemize}
\end{definition}

\begin{theorem}
        Let $f$ be a smooth function in a neighborhood 
    $N_x$ of $x=(x_1,\ldots,x_n)$ in $\mathbb{R}^n$. Suppose $f(0,\ldots,0)=0$. Then, 
    there exist $n$ smooth functions $g_i,\ldots,g_n$ defined on $N_x$ such that $g_i(0,\ldots,0)=\frac{\partial f}{\partial x_i}(0,\ldots,0)$
    for every $i$, and

    $$
        f(x_i,\ldots,x_n)=\sum_{i=1}^{n}{(x_i,\ldots,x_n)}
    $$
\end{theorem}
\begin{theorem}(Inverse Function Theorem)
        Let $f:\mathbb{R}^n\to \mathbb{R}^n$ be a smooth function on an open set $U$
        containing $a\in\mathbb{R}^n$. Suppose that $\det J_f(a)\not = 0$.
        \\Then there is an open set $V \subset \mathbb{R}^n$ containing $a$ and an open
        set $W\subset \mathbb{R}^n$ containing $f(a)$ such that $f:V\to W$ is a 
        diffeomorphism. 
\end{theorem}
\begin{theorem}(Morse Lemma.)
        Suppose that the point $a \in \mathbb{R}^k$ is a nondegenerate
        critical point of the function $f$, and
        $$
            (h_{ij})= \left(  \frac{\partial^2f}{\partial x_i \partial x_j} (a) \right)
        $$
        is the Hessian of $f$ at $a$. Then there exists a local coordinate system $(x_1,\ldots,x_k)$ around $a$ such that
        $$
            f=f(a)+\sum{h_{ij}x_ix_j}
        $$
        near $a$.
        \\
        or
        $$
            f=-x_1^2-x_2^2-\ldots-x_\lambda^2+x_\lambda^2+\ldots+x_n^2+f(a)
        $$
        where $\lambda$ is the index of $f$ at $a$.
\end{theorem}
\begin{proof}
     Let $p_0$ be a nondegenerate critical point of the function
     $f:\ M \to \mathbb{R}$, where $M$ is an $n$-manifold. The degeneracy of the point
     $p_0$ on $f$ is determined independent of our choice of a local coordinate system.
     Therefore, we may assume that when we pick a local coordinate system $(x_1,\ldots,x_n)$ defined
     in a neighborhood $N_{p_0}$,

@@

     When we square $y_1$, we get
     \begin{equation}
        y_1^2=\pm h_{11}x_1^2\pm2\sum_{i=2}^{n}{x_1x_ih_{1i}}\pm\frac{\left(\sum_{i=2}^{n}{x_ih_{1i}}\right)^2}{h_{11}}\tag{1.9}
     \end{equation}
     where the signs are either positive or negative, depending on the sign of $h_{11}$. Using equation
     $1.5$, we can verify that $f$ can be expressed in the following way with respect to this
     coordinate system on the restricted domain $\tilde{N}_0$.
     \begin{equation}
        f=\pm y_1^2 + \sum_{i=2}^{n}\sum_{j=2}^{n}{x_ix_jh_{ij}}-\frac{\left(\sum_{i=2}^{n}{x_ih_{1i}}\right)^2}{h_{11}}\tag{1.10}
     \end{equation}
     where the sign of the $y_1^2$ term is positive or negative, depending on the sign of $h_{11}$. Staying
     consistent with our goals, we notice that the first term is in the standard quadratic form seen
     in the Morse Lemma formulation, whereas the rest of the terms depend on local coordinates
     $x_i$ whereby $i\not=k \ (k=1)$. By induction from $k=1$ to $k=n$, we prove the Morse Lemma.
\end{proof}
% % end N1
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% % begin M1

\subsection{Examples of Morse functions} 
\begin{problem}
Show that the function $f: S^2 \to \R, f(x,y,z) = z$ is a Morse function.
\end{problem}
$f$ is smooth on $ S^2 $ since it extends to a smooth map on all of $\R^3$. We can map $R^3$ to $S^2$ with the stereographic projection two functions:
\begin{equation}
	\phi_1(x_1,x_2,x_3) = (\frac{x_1}{1-x_3}, \frac{x_2}{1-x_3} ) 
\end{equation}
and
\begin{equation}
	\phi_2(y_1,y_2,y_3) = (\frac{y_1}{1+y_3},\frac{y_2}{1+y_3})
\end{equation}
Therefore, to get $S^2 \to R^3$ we can take $\phi_1^{-1}$ and $\phi_2^{-1}$ \\
\begin{equation}
\phi_1^{-1}(x_1,x_2,x_3) = (\frac{2x_1}{x_1^2+x_2^2+1}, \frac{2x_2}{x_1^2+x_2^2+1}, \frac{x_1^2+x_2^2-1}{x_1^2+x_2^2+1})
\end{equation}
\begin{equation}
	\phi_2^{-1}(y_1,y_2,y_3) = (\frac{2y_1}{y_1^2+y_2^2+1}, \frac{2y_2}{y_1^2+y_2^2+1}, \frac{1-y_1^2+y_2^2}{y_1^2+y_2^2+1})
\end{equation}
Then, we take $g_1 = f \circ \phi_1^{-1}$ and $g_2 = f \circ \phi_2^{-1}$ \\
\begin{equation}
	g_1(x,y) =  \frac{x^2+y^2-1}{x^2+y^2+1}
\end{equation}
and
\begin{equation}
	g_2(x,y) = \frac{1-x^2+y^2}{x^2+y^2+1}
\end{equation}
Now, we can compute jacobian of $g_1$ and $g_2$, find critical points, and check that determinant of hessian matrix is non-zero. \\
$
	\nabla g_1(x,y,z) = 0$ iff $(x,y,z) = (0,0,-1)
$\\
$
	\nabla g_2(x,y,z) = 0$ iff $ (x,y,z) = (0,0,1)
$
We have two critical points, and now we compute the hessian at them.\\

\begin{equation}
	H(g_1) = \begin{bmatrix} 4 & 0 \\ 0 & 4 \end{bmatrix}
	\imply \det(H(g_1)) = 16
\end{equation}
\begin{equation}
	H(g_2) = \begin{bmatrix} -4 & 0 \\ 0 & -4 \end{bmatrix}
	\imply \det(H(g_2)) =  16
\end{equation}
Both critical points are non-degenerate, therefore $f$ is a Morse function.$\blacksquare$\\
% % end M1
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% % start V1
\\
\\
\begin{example}
$f(x,y)=e^{xy}+x$ is a Morse function
\end{example}
First of all we have to find all critical points of $f$. A point $P$ is critical, when $\frac{\partial f}{\partial x}=\frac{\partial f}{\partial y}=0$ at $P$. \\
$\frac{\partial f}{\partial x}= ye^{xy}+1=0 \\
\frac{\partial f}{\partial y}=xe^{xy}=0$
After solving the system of equations we get that $x=0$, $y=-1$. Hence, the only critical point of $f$ is $(0,-1)$. Now to prove that the function is a Morse function we have to shoe that the determinant of Hessian matrix at the point $(0,-1)$ is non-zero. \\
Partial second order derivatives are: \\
$\frac{\partial^2 f}{\partial x^2}=y^2e^{xy}$
$\frac{\partial^2 f}{\partial x^2}(0,-1)=1$\\
$\frac{\partial^2 f}{\partial \partial y}=e^{xy}+yxe^{xy}$
$\frac{\partial^2 f}{\partial \partial y}(0,-1)=1$\\
$\frac{\partial^2 f}{\partial y \partial x}=e^{xy}+xye^{xy}$
$\frac{\partial^2 f}{\partial y \partial x}(0,-1)=1$\\
$\frac{\partial^2 f}{\partial y^2}=x^2e^{xy}$
$\frac{\partial^2 f}{\partial y^2}(0,-1)=0$\\
Therefore, \\
$det(Hess(f(0,-1)))=\left| \begin{array}{cc} 1 & 1 \\ 1 & 0 \end{array} \right|=-1$. \\
Since the determinant is non-zero at the only critical point of the function, we can claim that it is a Morse function. \\
\begin{example}
Let's consider a function $g(x,y)=xy$ and prove that it is , indeed, a Morse function.\\

\end{example}
First order partial derivatives: \\
$\frac{\partial g}{\partial x}=y$\\
$\frac{\partial g}{\partial y}=x$\\
Therefore, the only critical point of $g$ is $(0,0)$. Now we have to evaluate the determinant of Hessian matrix at this point.\\
Second order partial derivatives: \\
$\frac{\partial^2 g}{\partial x^2}=0$\\
$\frac{\partial^2 g}{\partial x \partial y}=1$\\
$\frac{\partial^2 g}{\partial y \partial x}=0$\\
$\frac{\partial^2 g}{\partial^2 y}=1$\\
Hence the determinant of Hessian at the critical point is: \\
$det(Hess(f(0,0)))=\left| \begin{array}{cc} 0 & 1 \\ 1 & 0 \end{array} \right|=-1$. \\
We observe that the determinant is nonzero, hence, the only critical point of $g(x,y)=xy$ is non-degenerate, so the function is Morse. \\
\\
\\
\subsection{Examples of non-Morse functions} 
\begin{example}
$f(x,y)=x^3+xY^2-x^2y-y^3$
\end{example}
Let's follow the previous procedure: \\
$\frac{\partial f}{\partial x}=3x^2+y^2-2xy=0$\\
$\frac{\partial f}{\partial y}=2xy-x^2-3y^2=0$\\
After solving the system of equations we got that the only critical point of $f$ is $(0,0)$. Now compute second order partial derivatives:\\
$\frac{\partial^2 f}{\partial x^2}=6x-2y$\\
$\frac{\partial^2 f}{\partial x \partial y}=2y-2x$\\
$\frac{\partial^2 f}{\partial y \partial x}=2y-2x$\\
$\frac{\partial^2 f}{\partial^2 y}=2x-6y$\\
Therefore, \\
$det(Hess(f(0,0)))=\left| \begin{array}{cc} 0 & 0 \\ 0 & 0 \end{array} \right|=0$.\\
Hence, the only critical point of $f$ is degenerate, so $f$ is not a Morse function. \\
$g(x,y)=x^3$\\
$\frac{\partial g}{\partial x}=3x^2$\\
$\frac{\partial g}{\partial y}=0$\\
Therefore, critical points are all points of the form $(0, y_1)$.\\
$\frac{\partial^2 g}{\partial x^2}=6x$
$\frac{\partial^2 g}{\partial x \partial y}=0$\\
$\frac{\partial^2 g}{\partial y \partial x}=0$
$\frac{\partial^2 g}{\partial^2 y}=0$\\
Hence, Hessian will be of the form: \\
$Hess(g(0,y_1))=\left| \begin{array}{cc} 0 & 0 \\ 0 & 0 \end{array} \right|,  \forall y_1$. \\
Therefore, all critical points of $g$ are degenerate points, moreover, values of $g$ at all critical points are equal, hence, $g(x,y)=x^3$ is not a Morse function. 
% % end V1
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\section{Reeb graphs}
     \subsection{Introduction to Reeb graphs}
     \paragraph{}
      Suppose we have an equivalence relation, $\sim$, defined on a topological space $\X$. Let $\X_\sim$
      be the set of equivalence classes and let $\phi : \X \to \X_\sim$ map each point $x$ to its equivalence class.

\begin{definition}
      Let $f: \X \to \R$ be continuous and call a component of a level set a contour. Two
      points $x,y \in \X$ are equivalent if they belong to the same component of $f^{-1}(t)$
      with $t=f(x)=f(y)$. The Reeb graph of $f$, denoted as $R(f)=\X_\sim$, is the quotient space 
      defined by this equivalence relation.
  \end{definition}
      By construction, the Reeb graph has a point for each contour and the connection is 
      provided by the map $\phi:\X\to R(f)$. Letting $\pi:R(f)\to\R$ be defined such that $f(x)=\pi(\phi(x))$
      we can construct the level set by going backward, from the real line to the Reeb graph to the 
      topological space. Given $t\in\R$ we get $\pi^{-1}(t)$, a collection of points in $R(f)$,
      and $\phi^{-1}(\pi^{-1}(t))$, the corresponding collection of contours that make up the level set defined by $t$.

@@

\section{Applications}
\subsection{Topological Data Analysis}
A computational method for extracting simple descriptions of high dimensional data sets is a Mapper algorithm:
Let $X$ be a point cloud (data set) and we have a function $f:X \to \R$.
Assume that we have $N$ points in $X$ whose outputs are known for the function $f$.
Also, we assume that it is possible to compute inter-point distances between the points in the data.
Let $I$ be the range of the function $f$ restricted to the $N$ points and let $S$ be the covering of $I$.
We begin by dividing $S$ into a smaller intervals $(S_i)$ which overlap.
This gives us two parameters $l_i$ the length of $S_i$ and $p_{ij}$ the perceptange overlap between successive intervals.
Now for each $S_i$ we find a set $X_j = \{x | f(x) \in S_j\}$ s.t. $X_j \subset X$. The set $\{X_j\}$ forms a cover of $X$, and $X \subset \cup_j X_j$.
For each smaller set $X_j$ we find clusters $\{X_{jk}\} $.
We treat each cluster as a vertex in our reeb graph and draw an edge between vertices whenever $X_{jk} \cap X_{lm} \neq \emptyset$ i.e. the clusters corresponding to vertices have non-empty intersection. \\

@@

We have implemented the Mapper algorithm in python and ran it for the sklearn diabetes dataset and standford cars dataset, output are as follows:

\begin{center}
    diabetes 
    \includegraphics[width=.4\textwidth]{diabetes.png} \includegraphics[width=.4\textwidth]{cars.png}
    cars
\end{center}
Based on the diabetes data we may be able to make a difference between patients who don't need a treatmant, patients with diabetis type 1 and patients with diabetis type 2. While in the cars dataset it looks like there aren't any great differences between car features. (the graph is connected)

\subsection{Computer Graphics}
Computer graphics, in particular 3D modeling and animation, are a vast field. Building 3D models and analyzing their features are important both in entertainment (cartoons, animated films, computer games) and for scientific research (building 3D models based on photographs of a microscope for biology and medicine; building physical interactions in mechanics, astronomy, atomic physics ). There are two problems associated with 3D modeling, which are solved using Reeb graphs.\\

@@

\begin{thebibliography}{9}
     \bibitem{MorseTheory}
     {\sc Hua Amy}
       An introductory treatment of morse theory on manifolds
       \href{http://www.math.uchicago.edu/~may/VIGRE/VIGRE2010/REUPapers/Hua.pdf}{http://www.math.uchicago.edu/~may/VIGRE/VIGRE2010/REUPapers/Hua.pdf}
     \bibitem{Reebgraphs}
     {\sc Cole-McLaughlin at al.} 
       `` Loops in Reeb Graphs of 2-Manifolds  \href{https://core.ac.uk/download/pdf/20670562.pdf}{https://core.ac.uk/download/pdf/20670562.pdf}
     \bibitem{Mapper}
         {\sc Singh at al. }
          Loops in Reeb Graphs of 2-Manifolds \href{http://www.math.uchicago.edu/~may/VIGRE/VIGRE2010/REUPapers/Hua.pdf}{http://www.math.uchicago.edu/~may/VIGRE/VIGRE2010/REUPapers/Hua.pdf}
    \bibitem{Textbook}
        {\sc Allan Pollack and Victor Guillemin } 
        Differential Topology
\end{thebibliography}
\end{document}
